Graphical Tests for Power Comparison of Competing Designs

### 2012-03-31 experiment.tex
restructured a bit, filled in missing pieces

% !TEX root = comparison.tex
\section{Results and Discussion}~\label{results}

\subsection{Wind Direction and Airport Efficiency}
Figure \ref{fig:time} shows time taken until a decision was made in histograms, facetted by design and colored by correctness of answer. Because of the skewness of the distributions, times were log-transformed. Polar charts take on average more time to answer, but are answered with much lower precision. Figure \ref{fig:conf} shows two barcharts of confidence levels by task, again coloring is used for correctness of answers. Euclidean displays lead to a very bimodal distribution of confidence: participants are either very sure or not sure at all of their answer. Confidence levels in polar charts are distributed much more uniformly.

The perceptual tasks involved in decoding the designs fit into the general framework of the piechart versus barchart discussion. However, we are not dealing with assessing angles, which is known to be a perceptual harder task than assessments of heights or widths in bars, see e.g. \cite{cleveland:1984, robbins:2004, Kosslyn:2006, few:2009}. Instead, the perceptual tasks consists of comparisons along a common axis (in the barcharts) and a common origin (in the polar charts). The difference in designs is therefore based on how well we are able to judge deviations from a horizontal line compared to deviations from a circle. Based on this, the charts with added reference lines provide us with exactly the frame we compare with and should, therefore, be the `better' designs - either in speed or accuracy.

@@

\begin{figure}[hbtp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=\linewidth]{conf-answer} 
   \caption{Barcharts of (subjective) confidence in answering correctly for each lineup. A strongly bimodal distribution is apparent, while confidence levels for polar charts are more uniform. There are only slight (and non significant) difference between confidence levels and correctness of results.}
   \label{fig:conf}
\end{figure}

@@

For evaluating the results, we are using a generalized linear mixed effects modeling  approach \cite{pinheiro:2000} for all three of our response values, using the {\t R} package {\tt lme4} \cite{bates:2011}. 
For power predictions (cf. table \ref{tbl:correct}) a logistic regression was fitted for the competing designs, including covariates {\it sample size} (2, 4,6, 8, 10, and 24 percent of the data) and {\it shift} in wind direction (offset of 0, 90, 180, and 270 degrees), both as main effects and in two-way interaction effects with design to assess their effect on the power of designs. To adjust for individuals' ability, a random intercept was included in the model.

@@

Overall, the results show huge differences in the power of designs between polar charts and Euclidean: Euclidean designs are significantly more powerful than polar charts, particularly so with small sample sizes. 76.92\% of the Euclidean lineups shown resulted in the a correct identification of the real data while only 20.31\% of the polar charts were correct.
 The reference line has surprisingly little influence, but it helps more for  polar charts than for Euclidean charts,
An increase in sample size has positive impact on power (see figure \ref{fig:treatment} for an overview). Polar charts need a much bigger sample size to see an increase in power -- only at about 24\% of the original data do we see about the same power as for Euclidean charts of a sample size of 2\%.
The  changes in offset are significant -- interestingly, borderline behavior (90 and 270 degrees) does not show a difference between polar and Euclidean charts, whereas an inversion of the wave pattern (first up, then down), does show a difference. The power of Euclidean lineups suffers significantly  from this inversion whereas power of polar charts is unaffected. This is an unexpected finding, but is consistent throughout different lineups in the data. Figure \ref{fig:power} summarizes the results from the model: power predictions ($y$ axis) are shown by sample size ($x$ axis). The thick lines show average power by design for different shifts in wind direction. The thin lines represent power for individuals. What can be seen is the different impact of the offset by design: while  an offset of 270 degrees (the `mountain' pattern) has the highest power in Euclidean charts, it comes out worst in polar charts.

@@

Time taken to answer was log transformed before the modeling process to de-emphasize the impact of very large values (up to 500 seconds). The findings are consistent with correctness - time taken shows big differences between polar and Euclidean charts. The average amount of time spent on a Euclidean lineup is $e^{3.53} = 34.1$ seconds compared to $e^{4.07}=58.5$ seconds for a polar lineup.
The reference lines seem to increase the evaluation time, but not significantly. An increase in sample size decreases evaluation time, The inversion of the wave pattern at 180 degrees leads to a significant increase in time for Euclidean charts, but not for polar charts (at least not significantly).

>>>>

% !TEX root = comparison.tex
\section{Results and Discussion}~\label{results}
%
\subsection{Wind Direction and Airport Efficiency}
Figures \ref{fig:time} and \ref{fig:conf} give an overview of the relationship between the three response variables: accuracy, speed, and confidence level.

Figure \ref{fig:time} shows histogram of the time participants needed to make a decision on each lineup. Correctness of answers is shown by color.  Because of the skewness of the distributions, times were log-transformed. Polar charts take on average more time to answer, but are answered with much lower accuracy. 
The average amount of time spent on a Euclidean lineup is $e^{3.53} = 34.1$ seconds compared to $e^{4.07}=58.5$ seconds for a polar lineup.
This is in stark contrast to accuracy: 76.92\% of the Euclidean lineups shown resulted in the a correct identification of the real data while only 20.31\% of the polar charts were correct.

@@

Figure \ref{fig:conf} shows two barcharts of confidence levels by task, again coloring is used for correctness of answers. Euclidean displays lead to a very bimodal distribution of confidence: participants are either very sure or not sure at all of their answer. Confidence levels in polar charts are distributed much more uniformly.
\begin{figure}[hbtp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=\linewidth]{conf-answer} 
   \caption{Barcharts of (subjective) confidence in answering correctly for each lineup. A strongly bimodal distribution is apparent, while confidence levels for polar charts are more uniform. There are only slight (and non significant) difference between confidence levels and correctness of results.}
   \label{fig:conf}
\end{figure}

The perceptual tasks involved in decoding the designs fit into the general framework of the piechart versus barchart discussion. However, we are not dealing with assessing angles, which is known to be a perceptual harder task than assessments of heights or widths in bars, see e.g. \cite{cleveland:1984, robbins:2004, Kosslyn:2006, few:2009}. Instead, the perceptual tasks consist of comparisons along a common axis (in the barcharts) and a common origin (in the polar charts). The difference in designs is therefore based on how well we are able to judge deviations from a horizontal line compared to deviations from a circle. Based on this, the charts with added reference lines provide us with exactly the frame we compare with and should, therefore, be the `better' designs - either in speed or accuracy.

@@

For evaluating the results, we are using a generalized linear mixed effects modeling  approach \cite{pinheiro:2000} for each one of the three response values, using the {\t R} package {\tt lme4} \cite{bates:2011}. 
For power predictions (cf. table \ref{tbl:correct}) a logistic regression was fitted for the competing designs, including covariates {\it sample size} (2, 4,6, 8, 10, and 24 percent of the data) and {\it shift} in wind direction (offset of 0, 90, 180, and 270 degrees), both as main effects and in two-way interaction effects with design to assess their effect on the power of designs. To adjust for individuals' ability, a random intercept was included in the model.

@@

Overall, the results show huge differences in the power of designs between polar charts and Euclidean: Euclidean designs are significantly more powerful than polar charts, particularly so with small sample sizes. 
 The reference line has surprisingly little influence, but it helps more for  polar charts than for Euclidean charts,
An increase in sample size has a positive impact on power (see figure \ref{fig:treatment} for an overview). Polar charts need a much bigger sample size to see an increase in power -- only at about 24\% of the original data do we see about the same power as for Euclidean charts of a sample size of 2\%.
The  changes in offset are significant -- interestingly, borderline behavior (90 and 270 degrees) does not show a difference between polar and Euclidean charts, whereas an inversion of the wave pattern (first up, then down), does show a difference. The power of Euclidean lineups suffers significantly  from this inversion whereas power of polar charts is unaffected. This is an unexpected finding, but is consistent throughout different lineups in the data. Figure \ref{fig:power} summarizes the results from the model: power predictions ($y$ axis) are shown by sample size ($x$ axis). The thick lines show average power by design for different shifts in wind direction. The thin lines represent power for individuals. What can be seen is the different impact of the offset by design: while  an offset of 270 degrees (the `mountain' pattern) has the highest power in Euclidean charts, it comes out worst in polar charts.

@@

\begin{figure*}[hbtp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=0.9\linewidth]{power-exp2} 
   \caption{Overview of power predictions for the four different designs. The fully saturated thick lines show average predicted power for each of the designs facetted by size of the red group (top to bottom) and relative size of the blue group to the red group (left to right). Thin lines represent variability due to subject-specific abilities. }
   \label{fig:power2}
\end{figure*}

Time taken to answer was log transformed before the modeling process to de-emphasize the impact of very large values (up to 500 seconds). The findings are consistent with correctness - time taken shows big differences between polar and Euclidean charts. 
The reference lines seem to increase the evaluation time, but not significantly. An increase in sample size decreases evaluation time, The inversion of the wave pattern at 180 degrees leads to a significant increase in time for Euclidean charts, but not for polar charts (at least not significantly). Table \ref{tbl:time} shows an overview of the model parameters and estimates.

### 2012-03-31 setup.tex
restructured a bit, filled in missing pieces

@@

*** Null hypothesis for the lineup?
$Ho: $ wind direction has no effect on efficiency and $H_a:$ wind direction does affect efficiency.

@@

@@

 This results in a shift of the wave in Euclidean charts and a rotation in the polar charts as can be seen in figure \ref{fig:offset}. Our initial hypothesis was that this would have no effect on the polar charts, but might have a deteriorating effect on the Euclidean charts, in the case that the peak or the valley of the wave was at either end of the $x$ axis, i.e. an offset of 90 or 270 degrees. 
While these simple perturbations and different sample sizes allow us to get insight into different aspects of the designs, they also allow us to get multiple responses from each observer to assess individual ability without the need to go outside the framework of the original data, i.e. we leave the joint relationship between $x$ and $y$ essentially unchanged. For all of these combinations we produced two replicates, resulting in a total of 192 different lineups ($ 4 \text{ designs} \times 6 \text{ sample sizes} \times 4 \text{ offsets} \times 2 \text{ reps})$.

Data was collected using Amazon's Mechanical Turk (MTurk) service. XXX More details: website and screen shot.
 Each participant was shown ten lineups and asked to choose the graphic on each chart which they think is ``different''. They then identify, from a list of choices, why they chose this particular chart as well as how confident they are on a scale of one to five. Personal information such as age group, gender, and education is collected on a voluntary basis. 
The amount of time it took the individual to answer is also recorded. This  allows us to assess how design choices affect decoding \citet{cleveland:1994} in terms of accuracy and speed.

 In order to avoid people `gaming the system' as found in earlier studies \cite{heer:2010, kosara:2010}, participants were shown two specially prepared lineups as a baseline of performance -- if participants answered just one of these two reference lineups correctly, we required at least an overall correct percentage of 20\%, which is indicative of a performance that is better than guessing. Table \ref{tbl:treatment} gives an overview of the number of times lineups in each combination of design and sample size were shown, and in how many of them the data plot was correctly identified.

\begin{table}[hbtp]
\resizebox{\linewidth}{!} {
	%\rowcolors{1}{white}{lightgray}
	\begin{tabular}{ll@{   }r@{   }r@{   }r@{   }r@{   }r@{   }r}
	& & \multicolumn{6}{c}{sample size}  \\
        \multicolumn{2}{c}{type of chart} & 2 & 4 & 6 & 8 & 10 & 24 \\ [1pt] \hline 
	euclid & with & 12/{\bf 24}& 18/{\bf 27} & 42/{\bf 44} & 33/{\bf 41} & 39/{\bf 43} & 46/{\bf 47}\\
	       &      &(0.50)& (0.67) & (0.95) & (0.80) & (0.91) & (0.98)\\[1pt]
	& without & 13/{\bf 41} &24/{\bf 32}& 39/{\bf 44} & 47/{\bf 51} & 40/{\bf 45} & 34/{\bf 37}\\
	&         & (0.32) &(0.75)& (0.89) & (0.92) & (0.89) & (0.92)\\[3pt] %\hline
	polar & with & 13/{\bf 49}& 9/{\bf 43} & 9/{\bf 39} & 7/{\bf 35} & 8/{\bf 40} & 13/{\bf 38} \\
	      &      & (0.27)& (0.21) & (0.23) & (0.20) & (0.20) & (0.34) \\[1pt]
	& without & 6/{\bf 51}&   4/{\bf 34} &  5/{\bf 34} &  6/{\bf 39} & 11/{\bf 43} &  12/{\bf 37} \\
        &         & (0.12) & (0.12) &  (0.15) &  (0.15) & (0.26) &  (0.32)\\ 
	\end{tabular}
	}
\caption{\label{tbl:treatment} Breakdown of lineups: number correct/\textbf{number shown} (proportion correct). Each participant was shown eight lineups, and the same lineup was shown to multiple people (denominator in the table) reasonably well-spread among the treatment levels. }
\end{table}

@@

In a second experiment, we conducted a simulation study to investigate the power of four competing designs, as shown in figure\ref{fig:expii}, in assessing a mean shift between distributions. The null hypothesis of each lineup is $H_o:$ centers of the two groups are the same, and the alternative hypothesis is $H_a:$ center of the blue group is shifted to the right.

As covariates for this experiment, we considered the {\it size of the shift}, $d$, between the means of the distributions, with $d=0.4, 0.6, 0.8. 1.0,$ and $1.2$, as well as the sample size.  Since we are dealing with two groups of points, we varied both the size of the larger group, $n_1 = 15, 45$, and 135, as well as the relative size of the second group, with $n_2 = 1/3, 2/3$ and $3/3$ of $n_1$. For each of these combinations, we created three replicate data sets and four lineup charts, one for each of the competing designs. This resulted in 540 different lineup charts overall, which we sorted by $p$-value corresponding to the difference in means between the two groups in the simulated data. According to the $p$-value, we associated a `difficulty' level with lineups from 1 to 9 (easy to hard). Ten lineup plots were shown to each participant - one out of each of the difficulty levels and one additional reference chart with a particularly easy lineup  to allows for data quality checks.
\begin{figure} [hbtp]
   \centering
   \includegraphics[width=0.45\linewidth]{hist-conc} 
   \includegraphics[width=0.45 \linewidth]{boxplot-conc} \\
   \includegraphics[width=0.45 \linewidth]{density-conc} 
   \includegraphics[width=0.45 \linewidth]{dotplot-conc} 
   \caption{Overview of all four competing designs for displaying differences in distributions.}
   \label{fig:expii}
\end{figure}

>>>>

\subsection{Data Collection}
Data for both studies was collected using Amazon's Mechanical Turk (MTurk) service. 
The website for the second study is available at \url{http://www.public.iastate.edu/~mahbub/feedback_turk5/homepage.html}. Figure \ref{fig:screen} shows a screen shot of the website's layout for a dotplot lineup.

\begin{figure}[htbp] %  figure placement: here, top, bottom, or page
   \centering
   \includegraphics[width=\linewidth]{website} 
   \caption{Screenshot of the website for the second study. }
   \label{fig:screen}
\end{figure}

 Each participant is shown ten lineups and asked to identify the plot that is the most different from the other plots. In the second study the `difference' is further specified as `In which plot is the blue group the furthest to the right?". 
 In both studies, though, the question is placed prominently above each lineup as a reminder of what participants are asked to look for. 
  Additionally, participants are asked for a reason for their choice. They can select from a list of choices particular to the lineup design, why they chose this particular plot as well as how confident they are on a scale of one to five. 
 Personal information such as age group, gender, and education is collected on a voluntary basis. 
The amount of time it takes the individual to answer is also recorded. 
This  allows us to assess how design choices affect decoding \citet{cleveland:1994} in terms of accuracy and speed.

 In order to avoid people `gaming the system' as found in earlier studies \cite{heer:2010, kosara:2010}, participants were shown specially prepared lineups as a baseline of performance -- if participants fail to answer some of these baseline lineups, we require at least two correct answers for the remaining plots, which is indicative of a better-than-guessing performance, as inclusion criterion for the data evaluation.

@@

@@

In this scenario, we are dealing with the null hypothesis $H_0: $ wind direction has no effect on efficiency and the alternative $H_a:$ wind direction does affect efficiency.

@@

 This results in a shift of the wave in Euclidean charts and a rotation in the polar charts as can be seen in figure \ref{fig:offset}. Our initial hypothesis was that this would have no effect on the polar charts, but might have a deteriorating effect on the Euclidean charts, in the case that the peak or the valley of the wave was at either end of the $x$ axis, i.e. an offset of 90 or 270 degrees. 
While these simple perturbations and different sample sizes allow us to get insight into different aspects of the designs, they also allow us to get multiple responses from each observer to assess individual ability without the need to go outside the framework of the original data, i.e. we leave the joint relationship between $x$ and $y$ essentially unchanged. For all of these combinations we produced two replicates, resulting in a total of 192 different lineups ($ 4 \text{ designs} \times 6 \text{ sample sizes} \times 4 \text{ offsets} \times 2 \text{ reps})$.
Table \ref{tbl:treatment} gives an overview of the number of times lineups in each combination of design and sample size were shown, and in how many of them the data plot was correctly identified.

\begin{table}[hbtp]
\centering
%\resizebox{\linewidth}{!} {
	%\rowcolors{1}{white}{lightgray}
	\begin{tabular}{ll@{   }r@{   }r@{   }r@{   }r@{   }r@{   }r}
	& & \multicolumn{6}{c}{sample size}  \\
        \multicolumn{2}{l}{type of chart} & 2 & 4 & 6 & 8 & 10 & 24 \\ [1pt] \hline 
	euclid & with & 12/{\bf 24}& 18/{\bf 27} & 42/{\bf 44} & 33/{\bf 41} & 39/{\bf 43} & 46/{\bf 47}\\
	       &      &(0.50)& (0.67) & (0.95) & (0.80) & (0.91) & (0.98)\\[1pt]
	& without & 13/{\bf 41} &24/{\bf 32}& 39/{\bf 44} & 47/{\bf 51} & 40/{\bf 45} & 34/{\bf 37}\\
	&         & (0.32) &(0.75)& (0.89) & (0.92) & (0.89) & (0.92)\\[3pt] %\hline
	polar & with & 13/{\bf 49}& 9/{\bf 43} & 9/{\bf 39} & 7/{\bf 35} & 8/{\bf 40} & 13/{\bf 38} \\
	      &      & (0.27)& (0.21) & (0.23) & (0.20) & (0.20) & (0.34) \\[1pt]
	& without & 6/{\bf 51}&   4/{\bf 34} &  5/{\bf 34} &  6/{\bf 39} & 11/{\bf 43} &  12/{\bf 37} \\
        &         & (0.12) & (0.12) &  (0.15) &  (0.15) & (0.26) &  (0.32)\\ 
	\end{tabular}
%	}
\caption{\label{tbl:treatment} Breakdown of lineups: number correct/\textbf{number shown} (proportion correct). Each participant was shown eight lineups, and the same lineup was shown to multiple people (denominator in the table) reasonably well-spread among the treatment levels. }
\end{table}

@@

In a second experiment, we conducted a simulation study to investigate the power of four competing designs, as shown in figure \ref{fig:expii}, in assessing a mean shift between distributions. The null hypothesis of each lineup is $H_o:$ centers of the two groups are the same, and the alternative hypothesis is $H_a:$ center of the blue group is shifted to the right.

As covariates for this experiment, we considered the {\it size of the shift}, $d$, between the means of the distributions, with $d=0.4, 0.6, 0.8. 1.0,$ and $1.2$, as well as the sample size.  Since we are dealing with two groups of points, we varied both the size of the larger group, $n_1 = 15, 45$, and 135, as well as the relative size of the second group, with $n_2 = 1/3, 2/3$ and $3/3$ of $n_1$. 
For each of these combinations, we created three replicate data sets with values from an Exponential distribution with rate parameter $\lambda=1$ (for the first group) and rate parameter $\lambda_2=1/(d+1)$, which corresponds to a mean shift of $d$, for the second groups. 
Based on each data set we created four lineup charts, one for each of the competing designs. This resulted in 540 different lineup charts overall, which we sorted by $p$-value corresponding to the difference in means between the two groups in the simulated data. According to the $p$-value, we associated a `difficulty' level with lineups from 1 to 9 (easy to hard). Ten lineup plots were shown to each participant - one out of each of the difficulty levels and one additional reference chart with a particularly easy lineup  to allow for data quality checks.
\begin{figure} [hbtp]
   \centering
   \includegraphics[width=0.45\linewidth]{hist-conc} 
   \includegraphics[width=0.45 \linewidth]{boxplot-conc} \\
   \includegraphics[width=0.45 \linewidth]{density-conc} 
   \includegraphics[width=0.45 \linewidth]{dotplot-conc} 
   \caption{Overview of all four competing designs for displaying differences in distributions.}
   \label{fig:expii}
\end{figure}
